---
title: "Habitat actions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Habitat actions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, fig.width = 6, fig.height =4)
```

```{r setup, echo = FALSE, message = FALSE, warning = FALSE}
library(R2Rscenario)
library(fallRunDSM)
library(DSMhabitat)
library(DSMflow)
```

This document provides additional information about how habitat actions are implemented in the fall run model for Reorienting to Recovery (R2R). For more more information about the model itself, including the codebase, please see [our documentation](https://reorienting-to-recovery.github.io/fallRunDSM/).

The habitat you select interacts with your flow selection, as habitat input objects are generated by relating flow to suitable area. For more information, please see [the DSMhabitat documentation](https://reorienting-to-recovery.github.io/DSMhabitat/) and the [DSMflow documentation](https://reorienting-to-recovery.github.io/DSMflow/).

## Habitat actions

There are two steps to building out the habitat inputs for the model. The first is to select your base habitat input. There are 4 options and they are mutually exclusive:

* `1`: Baseline Habitat
* `2`: Theoretical Max Habitat 
* `3`: Healthy Rivers and Landscapes (HRL) habitat

You can then layer on additional habitat actions:

* `4`: Add Rice Lands Salmon Rearing Practice Standard 
* `5`: Increase prey density 
* `6`: Decrease predation (scale contact points by 1/3)
* `7`: Add 20,000 acres of fish food production (HRL)
* `8`: Remove predation contact points (HRL)
* `27`: Add effects of spring run weir on fall run
* `28`: Add effects of above-dam spring run habitat on fall run
* `29`: Add San Joaquin floodplain habitat

### Base Habitat Input

**Implemented in all watersheds and years.**

#### 1: Baseline Habitat

This habitat action represents current habitat used in the SIT models as well as current and planned restoration projects. For further documentation on how project information was solicited and modeled, please see [this documentation](https://reorienting-to-recovery.gitbook.io/documentation-site/zCZ2Z2yqFYMUQrtZdTlg/modeling-documentation/scenarios/bookend-scenario-overview/baseline-overview/reorienting-to-recovery-baseline-habitat-methodology-and-overview). 

#### 2: Theoretical Max Habitat

This habitat action represents the total potential habitat available for salmonids, above- and below-dams in the system. For further documentation on how the theoretical maximum habitat values were calculated and incorporated into the model, please see [this documentation](https://reorienting-to-recovery.gitbook.io/documentation-site/zCZ2Z2yqFYMUQrtZdTlg/modeling-documentation/scenarios/bookend-scenario-overview/theoretical-maximum-habitat-overview). 

#### 3: HR&L Habitat

This habitat action was designed to approximate the effects of ongoing and proposed projects under the Healthy Rivers and Landscapes Program, as outlined in the [Memorandum of Understanding, Appendix 2](https://resources.ca.gov/-/media/CNRA-Website/Files/NewsRoom/email-items/VoluntaryAgreementMOUTermSheet20220329_SIGNED-20220811.pdf). Some actions in the MOU are available as "layered" on actions (i.e. actions `7`, `8`). For further documentation on how these actions were incorporated into the model, please see [this documentation](https://reorienting-to-recovery.gitbook.io/documentation-site/zCZ2Z2yqFYMUQrtZdTlg/modeling-documentation/scenarios/balanced-scenario-overview/elephant-overview/elephant-habitat-analysis). 

### Layered actions

#### 4: Add Rice Lands Salmon Rearing Practice Standard

This action models the effect of managing rice fields (wet and dry side, outside the bypasses) to increase floodplain habitat and prey density in Sacramento watersheds. This is modeled by increasing floodplain in the Yolo Bypass in January and February by a factor of `9000` acres.

**All years, Yolo Bypass**

#### 5: Increase prey density 

This action models the effect of improving habitat through food/prey subsidy management practices to increase salmon growth and survival. This is modeled by setting prey density in all watersheds (including the delta) and model years to "hi", the maximum level possible.

**All years, customizable watersheds**

#### 6: Decrease predation

This action models the effect of removing predator contact points to decrease predation on juvenile salmonids by decreasing the number of contact points in a watershed by a factor of `1/3`. Contact points are sourced from the [CalFish Predator Assessment Database (PAD)](https://www.calfish.org/ProgramsData/HabitatandBarriers/CaliforniaFishPassageAssessmentDatabase.aspx&sa=D&source=docs&ust=1731691522134452&usg=AOvVaw2plXmmET6Hihl1xQKNmxGn). 

**All years, all watersheds**

#### 7: 20,000 acres of fish food production prey density (HRL)

This action approximates the effect of adding `20,000` acres of fish food production as outlined in the [Memorandum of Understanding, Appendix 2](https://resources.ca.gov/-/media/CNRA-Website/Files/NewsRoom/email-items/VoluntaryAgreementMOUTermSheet20220329_SIGNED-20220811.pdf). This is modeled by setting prey density to "hi" in the Sutter Bypass and Sacramento River (to represent Colusa Basin) to coarsely represents effects of increasing 20,000 acres off-channel. Additionally, the available acreage for dry side floodplain for Sutter and Colusa basins was calculated using [these data](https://urldefense.proofpoint.com/v2/url?u=https-3A__flowwest.shinyapps.io_riceflows4ff_&d=DwMGaQ&c=euGZstcaTDllvimEN8b7jXrwqOf-v5A_CdpgnVfiiMM&r=es1mYD_t8ssn_HUUEGOgPpeBoWf2_Iri3BC8scPyjqg&m=MPH46l1pSkkFUhwOq2afwCu7FQy9x-3ynNLgjT1FrFVENIGYvlHv77LDW2IA1HMx&s=gks_3V6uaP9BPa4q9vmFCDSgHkke0BCTa3QXxOITYfw&e=). The proportion of the committed 20,000 acres those acreages provide was calculated and then used to scale instream habitat for those watersheds for the months of January and February. 

**All years, Sutter Bypass and Lower + Lower-mid Sacramento River**

#### 8: Remove predation contact points (HRL)

This action models a reduction in predator contact points based on communications with parties for proposed or current projects. This is currently modeled by removing `1` predator contact point for Feather River, `9` for the Upper Sacramento River, `1` for Antelope Creek, and `1` for Butte Creek.

**All years, Feather River, Upper Sacramento River, Antelope Creek, and Butte Creek**

#### 27: Effect of spring run weir

This action models the effect of installing a weir to protect spring run spawning habitat. This is modeled differently for tributaries where spring- and fall-run spawning habitat completely overlap (Upper Sacramento River, Antelope Creek, Feather River, Mokelumne River, Stanislaus River, and Tuolumne River), partial overlap (Big Chico Creek, Clear Creek, and Mill Creek), and for the Yuba River. For tributaries with full overlap, we reduce fall-run spawning habitat by `1/3`. For tributaries with partial overlap, we subtract the overlapping part from fall-run spawning habitat. For Yuba River, we reduce fall-run spawning habitat by half to approximate the effect of installing a weir at Daguerre Point Dam. 

**All years, spring-run watersheds - specific for fall run model**

#### 28: Effects of above-dam spring run habitat

This is modeled by reducing in-channel and floodplain rearing habitat for fall-run by a factor of `10%` (in development, may be iteratively improved) for tributaries where spring-run are present: Upper Sacramento River, Antelope Creek, Feather River, Mokelumne River, Stanislaus River, Tuolumne River, Big Chico Creek, Deer Creek, Mill Creek, Clear Creek, Yuba River, and Battle Creek. This is to approximate the effect of above-dam habitat producing more spring-run juveniles, which increases competition for space in rearing habitat for fall-run juveniles.

**All years, spring-run watersheds - specific for fall run model**

#### 29: Add San Joaquin floodplain habitat

This action models an increase in San Joaquin floodplain habitat by scaling existing floodplain habitat up to `50%` of the theoretical maximum habitat calculated for the area (see action `2`).

**All years, San Joaquin River**

